---
title: "Hashtag oligo barcode QC"
subtitle: "analyzing the HTO distribution for the IFNA-BM-Niche project"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    theme: paper
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: true
    fig_width: 7
    fig_height: 6
---

```{r setup, include=FALSE, error=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE,
                      include = TRUE,
                      warning = FALSE, 
                      message = FALSE, 
                      cache.lazy = FALSE, 
                      tidy = FALSE)

knitr::opts_chunk$set(options(stringsAsFactors = FALSE))
```
  
<br/><br/><br/>  

# Setting up the QC pipeline  

<br/><br/> 
  
## Packages  

<br/><br/>

```{r 1.1_Packages}

# choose packages to install and/or load
pkg <- c("knitr", "kableExtra",
         "dplyr", "tibble", "tidyr", "stringr", "magrittr",
         "ggplot2", "grid", "gridExtra", "RColorBrewer",
         "Matrix", "Seurat", "caret")

new.pkg <- pkg[!(pkg %in% installed.packages())]

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

if (length(new.pkg)) {
  BiocManager::install(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
} else sapply(pkg, require, character.only = TRUE)

options(stringsAsFactors = FALSE)
```

<br/><br/>

## input and output directories  

<br/><br/>

```{r 1.2-Directories}
data.dir <- "/data/dir"
```

<br/><br/>
  
# Loading data and curating dataset  
  
<br/>
  
The output of the CITE-Seq Count pipeline is a count matrix (reads and UMI counts), consisting of:  
a file with the **feature IDs**, representing all hashtags (HTOs) quantified,  
a file with the **cell IDs**, representing all cell barcodes quantified,  
a **matrix** of counts per hashtag for every cell.  
<br/>  
Plus it also provides a report.yaml, a list of uncorrected cells, and unmapped reads, if of interest.

<br/><br/>

## Loading the data
  
<br/>

```{r 2.1_ReadData}
# list all files in dir dir
Experiment <- list.files(path = data.dir, include.dirs = T)

# choose file for downstream QC
file <- 1

# read in count matrix  
HTO_m <- Read10X(file.path(data.dir, Experiment[file], "umi_count"), 
                 gene.column = 1)

# adapt cell bcs to have 10x standard
colnames(HTO_m) <- paste(colnames(HTO_m), "1", sep = "-")
#head(colnames(HTO_m))

# order by descending counts
counts_per_cell <- Matrix::colSums(HTO_m[1:4,])
HTO_m <- rbind(HTO_m, counts_per_cell)
HTO_m <- HTO_m[,order(HTO_m[4,], decreasing = T)]

# show table
HTO_m[,1:3] %>%
  kable(caption = "First three entries of HTO Count matrix") %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```
  
<br/>

For experiment   **`r Experiment[file]`**:  
Overall, `r ncol(HTO_m)` cell barcodes could be recovered.  
`r length(counts_per_cell[counts_per_cell==0])` cells have **0** HTO barcode count,  
`r length(counts_per_cell[counts_per_cell>0])` cells have **at least 1** HTO barcode count,  
`r length(counts_per_cell[counts_per_cell>10])` cells have **at least 10** HTO barcode counts.  
  
<br/>  

## Filtering the data

<br/>
  
For downstream QC and visualization, cells with 0 HTO counts will be removed.  
  
<br/>  
  
```{r 2.2_FilterData}

## filter HTO count matrix
# only keep cells that have at least 1 read
HTO_m <- HTO_m[,Matrix::colSums(HTO_m[1:4,]) >0]

HTO_df <- as.data.frame(t(HTO_m)) %>%
  arrange(desc(counts_per_cell)) %>%
  rowid_to_column("rank") %>%
  tidyr::gather(key = Tag, value = value, 
                `tag1-ACCCACCAGTAAGAC`:`tag4-AAAGCATTCTTCACG`) %>%
  arrange(rank) %>%
  filter(value > 0)

#head(HTO_df)

```
  
<br/><br/>
  
# Visualizing results 

<br/><br/>

### Count depth

<br/>

```{r 3.1-Plotting_count_depth, fig.height=6 ,fig.width=10}

def.par <- par()

## overall counts per cell
counts_per_cell <- Matrix::colSums(HTO_m[1:4,])

## show overall count depth distribution as ranksize
plot(sort(counts_per_cell, decreasing = TRUE),
     xlab = "Barcode rank", ylab = "counts per cell",
     main = "Barcode Rank plot - all counts", log="xy", col ="royalblue")

cat("\n")

## show count depth distribution per tag as ranksize
cols <- brewer.pal(n = 4, name = "Set1")
layout(matrix(c(1,2,3,4), 2, 2, byrow = TRUE))
for (i in 1:4) suppressWarnings({
plot(sort(HTO_m[i,], decreasing = TRUE),
             xlab = "Barcode rank", ylab = "counts per cell",
             main = paste("Count depth for", rownames(HTO_m)[i]), 
     log="xy", col =cols[i])
})
suppressWarnings(par(def.par))

cat("\n")

#plot counts per cell

p1 <- ggplot(data = HTO_df, aes(x=counts_per_cell, fill = Tag)) +
  geom_histogram(position = "stack",
                 #alpha=0.5,
                 binwidth = .1,
                 #col="black",
                 size=1) +
  scale_fill_brewer(palette = "Spectral") +
  scale_x_log10() +
  theme_minimal() + 
  xlab("log10 Count depth") + 
  #facet_grid(rows = "Tag") +
  ggtitle("Expression sum per hashtag per cell") +
  theme(legend.position = "none")

p2 <- ggplot(data = HTO_df %>% filter(counts_per_cell < 11),
       aes(x=counts_per_cell, fill = Tag)) +
  geom_histogram(binwidth = 1,
                 #col="black",
                 size=1) +
 # scale_x_log10()+
  scale_fill_brewer(palette = "Spectral") +
  theme_minimal() +
  xlab("Count depth") +
  ggtitle("Expression sum per hashtag per cell   < 11 counts") 

 p3 <- ggplot(data = HTO_df %>% filter(counts_per_cell < 100 & counts_per_cell > 10),
       aes(x=counts_per_cell, fill = Tag)) +
  geom_histogram(aes(fill=Tag),
                 binwidth = 1,
                 #col="black",
                 size=1) +
 # scale_x_log10()+
  scale_fill_brewer(palette = "Spectral") +
  theme_minimal() +
  xlab("Count depth") +
  ggtitle("Expression sum per hashtag per cell   11 - 99 counts") +
  theme(legend.position = "none")

p4 <- ggplot(data = HTO_df %>% filter(counts_per_cell > 99), 
      aes(x=counts_per_cell, fill = Tag)) +
  geom_histogram(aes(fill=Tag),
                 binwidth = 100,
                 #col="black",
                 size=1) +
  #scale_x_log10()+
  scale_fill_brewer(palette = "Spectral") +
  theme_minimal() +
  xlab("Count depth") +
  ggtitle("Expression sum per hashtag per cell   > 99 counts") +
  theme(legend.position = "none")

cowplot::plot_grid(p1, p2, p3, p4,
                   rel_widths = c(1, 1, 1, 1),
  ncol = 2) +
  cowplot::panel_border()

cat("\n")
suppressWarnings(par(def.par))
```

<br/>

## Overall hashtags detected

<br/>

```{r 3.2-Plotting_tags_per_cell, fig.height=6 ,fig.width=10}
# Let's check how many HTO tags can be detected per cell and
# investigate the distribution
## number of tags being detected per cell
tags_per_cell <- Matrix::colSums(HTO_m[1:4,]>0)

table(tags_per_cell) %>%
  kable(caption = "Total number of HTO tags per cell") %>%
   kable_styling(bootstrap_options = c("striped", "hover", 
                                       "condensed"),full_width = F)

cat("\n")

#plot tags per cell
t <- as.data.frame(table(tags_per_cell))
t$sqFreq <- sqrt(t$Freq)

def.par <- par()
par(mar=c(4,4,4,4))
barplot(t$Freq, main="Tags per cell", horiz = T,
        las=1, xlim=range(pretty(c(0, t$Freq))),
        xlab = "Number of cells",
        ylab = "Number of tags",
        names.arg = t$tags_per_cell,
        col="#69b3a2")

cat("\n")

par(mar=c(4,4,4,4))
barplot(t$sqFreq, main="Tags per cell", horiz = T,
        las=1, xlim=range(pretty(c(0, t$sqFreq))),
        xlab = "Number of cells (sqrt)",
        ylab = "Number of tags",
        names.arg = t$tags_per_cell,
        col="#69b3a2")

cat("\n")

suppressWarnings(par(def.par))
```

<br/>
  
##  Counts & Cells per hashtag

<br/>
  
```{r 3.3-Plotting_X_per_tag, fig.height=8 ,fig.width=12}

# Let's check how many HTO tags could be found in the library and
# how many cells have the hash tags
## overall number of HTO hashtags found
counts_per_tag <- Matrix::rowSums(HTO_m[1:4,])

stack(counts_per_tag) %>%
  arrange(desc(values)) %>%
  rename(count=values, HTO=ind) %>%
  kable(caption = "Total number of HTO tags being found (in decreasing order)") %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
  
cat("\n")

# plot counts per tag
def.par <- par()
par(mar=c(4,12,4,4))
barplot(counts_per_tag, main="Counts per tag", horiz = T,
        las=1, xlim=range(pretty(c(0, counts_per_tag))),
        col="#69b3a2")

cat("\n")


## number of cells having the specific HTO hashtag
cells_per_tag <- Matrix::rowSums(HTO_m[1:4,]>0) # only count cells where at least 1 tag is expressed

stack(cells_per_tag) %>%
  arrange(desc(values)) %>%
  rename(count=values, HTO=ind) %>%
  kable(caption = "Number of cells having the specific HTO hashtag (non-unique)") %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

cat("\n")

# plot cells per tag
par(mar=c(4,12,4,4))
barplot(cells_per_tag, main="Cells per tag (non-unique)", horiz = T,
        las=1, xlim=range(pretty(c(0, cells_per_tag))),
        col="#69b3a2")

par(def.par)
```

<br/>
<br/>

```{r 5.1-SessionInfo}
sessionInfo()
```
